---
title: "MARINe_CTI"
author: "Erica Nielsen"
date: "2023-11-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is code to take the 'phototranssummarysd' data from MARINe and calculate community temperature index (CTI) per site per year, and run statistical models to look at influence of environment on CTI change.


```{r import_data}
#read in data
phototran_df <- read.csv("C:/Users/erica.nielsen/Desktop/Synz/MARINe/large_data/phototransummarysd_download.csv")

#filter to only CA sites (can extend in future if we want)



```

```{r filter}
library(dplyr)
#retain only columns of interest
trans_filt<-phototran_df[,c(3,4,5,9,10,15,16,17,18,25)]

#sort by spp per site for each year
trans_by_species<- trans_filt %>% group_by(species_code, marine_common_year, site_code, site_lat)

#get avg % cover per spp per site per year
cov_per_yr<-trans_by_species %>% summarise(
    cov = mean(average_percent_cover)
)

#only keep rows with per cov greater than zero
cov_per_yr_filt<-cov_per_yr[(cov_per_yr[,5]>0),]


#get min/max lat per year each spp is found at
spp.min.max.lat<-cov_per_yr_filt %>% group_by(species_code, marine_common_year) %>%
        summarise(max_lat = max(site_lat), min_lat = min(site_lat))

```

```{r check temp}

#now we will check to see if MARINe temp data is avaiable for all sites
temp_means <- read.csv("C:/Users/erica.nielsen/Desktop/Synz/MARINe/PCI_MARINe_synth/data/MARINe_daily_temperature_means_UTC.csv")

matched <- intersect(cov_per_yr_filt$site_code, temp_means$marine_site_code)
all <-  union(cov_per_yr_filt$site_code, temp_means$marine_site_code)
non.matched <- all[!all %in% matched]

## There's 51 sites that are not in the temp data

# let's just see if we try to extract temp per year per lat what will happen...

#first remove unneccessary columns from temp data
temp_filt<-temp_means[,c(2,4,11,18)]

#take mean temp per year
temp_mean_yr<-temp_filt %>% group_by(marine_site_code, year) %>%
        summarise(temp = mean(mean))
# the above DF shows that many sites don't have temp data going back all years... might want to use something more consistent

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
