---
title: "MARINe_GDM"
author: "Erica Nielsen"
date: "2023-12-19"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Generalized Dissimarity Models Analyses for PCI

This markdown takes biodiversity data from MARINe rocky shore long term monitoring & satellite environmental data to run GDMs to map community structure for the present dat & 2050. 

```{r Read biodiversity data, warning=FALSE}
library(readxl)

#read point contact
pt_cont <- read_excel("C:/Users/erica.nielsen/Desktop/Synz/MARINe/large_biodiv_data/cbs_data_CA_2023.xlsx", 
    sheet = "point_contact_summary_data")

#read quadrat summary
quad <- read_excel("C:/Users/erica.nielsen/Desktop/Synz/MARINe/large_biodiv_data/cbs_data_CA_2023.xlsx", 
    sheet = "quadrat_summary_data")

#read swath data
swath <- read_excel("C:/Users/erica.nielsen/Desktop/Synz/MARINe/large_biodiv_data/cbs_data_CA_2023.xlsx", 
    sheet = "swath_summary_data")

```

The input data to make 'site-pair' tables for GDMs -> Rows = sites, Columns = spp abundance. To do this we need to do the following:

-combine pt_cont, quad, & swath tables into single table... problem -> quad & swath are density/m2 and pt_cont is % cov
  - we are just comparing sites, so if %cov and density 


```{r check spp per table, warning=FALSE}

#see if there's overlapping spp between three tables 
intersect(intersect(pt_cont$species_lump, quad$species_lump), swath$species_lump)
character(0) # so there's no species overlap between tables

```

The next problem is that we need one 'spp_abund' count for the present day and we have multiple abundances per species (per year)-- will try and average spp abundaces over the years that the environmental data is available for
(I checked the air temp data from MARINe but it doesn't exist for all biodiv sites so will need to use other data)

- WorldClim Version 2 = 1970-2000, so not the best temporal range to match data
--> looking into air temp for same range, and this seems like a good option:                    

    https://climatedataguide.ucar.edu/climate-data/airs-atmospheric-infrared-sounder-version-6-level-2 
        download here: https://disc.gsfc.nasa.gov/datasets?page=1&source=AQUA%20AIRS,AQUA%20AMSU-A,AQUA%20HSB
            date range=2002-09-01 to 2014-09-01
            spatial range= xmin= -126, xmax= -116, ymin = 32, ymax= 42


- Bio-Oracle version 2 = 2000-2014 



```{r filter_years, message=FALSE, warning=FALSE}


library(dplyr)
library(ggplot2)

#filter by year per DF

quad_yr<-quad%>% filter(between(year, 2000, 2014))
swath_yr<-swath%>% filter(between(year, 2000, 2014))
pt_cont_yr<-pt_cont%>% filter(between(year, 2000, 2014))

#now see how many years each site was sampled for
  # think I might want to filter to include only sites with 3+ years... or run analyses on all data vs. filtered data

## QUAD
yrs_quad<-quad_yr %>% group_by(marine_site_name,latitude) %>% count(year, sort = TRUE)

yrs_quad<-yrs_quad %>%
    group_by(marine_site_name) %>%
    add_count(name = "num_years")

ggplot(yrs_quad, aes(x=latitude, y=num_years)) + geom_point()
#a lot of the northern sites only have 1 year collected for quad


## SWATH
yrs_swath<-swath_yr %>% group_by(marine_site_name,latitude) %>% count(year, sort = TRUE)

yrs_swath<-yrs_swath %>%
    group_by(marine_site_name) %>%
    add_count(name = "num_years")

ggplot(yrs_swath, aes(x=latitude, y=num_years)) + geom_point()
# similar pattern in swath

## PT Cont
yrs_pt_cont<-pt_cont_yr %>% group_by(marine_site_name,latitude) %>% count(year, sort = TRUE)

yrs_pt_cont<-yrs_pt_cont %>%
    group_by(marine_site_name) %>%
    add_count(name = "num_years")

ggplot(yrs_pt_cont, aes(x=latitude, y=num_years)) + geom_point()
#even fewer years for the pt_cont data


```

-Looks like doing >3 years per site would yield very low sample sizes... think that we can try comparing 1 vs. 2 years sampling and see if there's difference

- What I could also do is not filter the data to 2000-2014, then apply 3+ year filter and see if outputs differ (only thing then is that the climate data misses 2015-2023)


```{r get avg abund, warning=FALSE, message=FALSE}
## Starting with no number of yrs filter

#average abundance per species over the years per site
quad_avg <- quad_yr %>%
     group_by(marine_site_name, latitude, longitude, species_lump) %>%
     summarize(Mean = mean(density_per_m2, na.rm=FALSE))

pt_cont_avg <- pt_cont_yr %>%
     group_by(marine_site_name, latitude, longitude, species_lump) %>%
     summarize(Mean = mean(percent_cover, na.rm=FALSE))

swath_avg <- swath_yr %>%
     group_by(marine_site_name, latitude, longitude, species_lump) %>%
     summarize(Mean = mean(density_per_m2, na.rm=FALSE))

#join data frames
common_col_names <- intersect(names(quad_avg), names(pt_cont_avg)) #get column names
quad_pt<-merge(quad_avg, pt_cont_avg, by=common_col_names, all.x=TRUE, all.y=TRUE) #merge first 2 DFs
quad_pt_swath<-merge(quad_pt, swath_avg, by=common_col_names, all.x=TRUE, all.y=TRUE) #merge in swath

```

So now we have a 'x-y species list' dataframe (which is bioFormat=2 for GDM, i.e. spp are rows and not columns)

Last step before we can make a 'site-pair' table is to get the environmental predictor variables

```{r get ocean vars, warning=FALSE}
library(sdmpredictors)


```






